An Empirical Study of Invasion and Resistance for Iterated Prisoner's Dilemma Strategies

### 2017-05-18 main.tex
Correct minor typos

\begin{enumerate}
 \item Zero determinant strategies are not particularly effective for $N > 2$
 \item Complex strategies can be effective, and in fact can naturally evolve
     through evolutionary processes to outperform designed strategies.
 \item Strong resistors specifically evolve or have a handshake mechanism.
 \item Strong invaders are generally cooperative strategies that do not defect
 first but retaliate to varying degrees of intensity against strategies that defect.
 \item Strategies evolved to maximize their total payout can be strong invaders
 and achieve mutual cooperation with many other strategies.
\end{enumerate}

@@

Figure~\ref{fig:moran_process} shows a diagrammatic representation of the Moran
process, a stochastic birth death process on a finite population in which the
population size stays constant over time. Individuals are \textbf{selected}
according to a given fitness landscape. Once selected, the individual is
reproduced and similarly another individual is chosen to be removed from the
population. In some settings mutation is also considered but without mutation
(the case considered in this work) this process will arrive at an absorbing
state where the population is entirely made up of players of one strategy. The
probability with which a given strategy takes over a the population is called the
\textit{fixation probability}. A more detailed analytic description of this is
given in Section~\ref{sec:validation}. In our simulations offspring do not
inherit any knowledge or history from parent replicants.

@@

The Moran process was initially introduced in \cite{Moran1957}. It has since
been used in a variety of settings including the understanding of the spread of
cooperative and non-cooperative behaviour such as cancer \cite{West2016} and the
emergence of cooperative behaviour in spatial topologies \cite{Nowak2017}.
However these works mainly consider non-sophisticated strategies. Some work has
looked at evolutionary stability of strategies within the Prisoner's Dilemma
\cite{Li2014} but this is not done in the more widely used setting of the Moran
process, rather in terms of infinite population stability. In \cite{Baek2016}
Moran processes are studied in a theoretical framework for a small subset of
strategies. The subset included memory one strategies, strategies that recall
the events of the previous round only.

Of particular interest are the Zero determinant strategies introduced in
\cite{Press2012}. It was argued in \cite{stewart2013extortion} that generous
ZD strategies are robust against invading strategies. However, in \cite{Lee2015}
a strategy using machine learning techniques was capable of resisting invasion
and also able to invade any memory one strategy. Recent work \cite{Hilbe2017}
has investigated the effect of memory length on strategy performance and the
emergence of cooperation but this is not done in a Moran process context and only
considers specific cases of memory 2 strategies. In \cite{Adami2013} it was
recognised that many zero determinant strategies do not fare well against
themselves. This is a disadvantage for the Moran process where the best
strategies cooperate well with other players using the same strategy.

@@

To carry out this large numerical experiment, \input{./tex/num_strategies}
strategies, listed (with their properties) in Appendix~\ref{app:list_of_players}
are used from the Axelrod library. There are
\input{./tex/num_stochastic.tex}stochastic and
\input{./tex/num_deterministic.tex}deterministic strategies. Their memory depth,
defined by the number of rounds of history used by the strategy each round, is
shown in Table~\ref{tbl:memory_depth_count}. The memory depth is infinite if the
strategy uses the entire history of play (whatever its length). For example, a
strategy that utilizes a handshaking mechanism where the opponents actions on
the first few rounds of play determines the strategies subsequent behavior would
have infinite memory depth.

@@

A part from the PSO Gambler strategy, which was trained using a particle swarm
optimisation algorithm, these strategies are trained with an evolutionary
algorithm that perturbs strategy parameters and optimizes the mean total score
against all other opponents \cite{affenzeller2009genetic}. They were trained to
win IPD tournaments by maximizing their mean total payoffs against a variety
of opponents. Variation is
introduced via mutation and crossover of parameters, and the best performing
strategies are carried to the next generation along with new variants. Similar
methods appear in the literature \cite{Ashlock2006}.

@@

Each matchup of players was run to fixation for the specified number of
repetitions to estimate the absorption probabilities. The trained algorithms
were run for less than 50 generations. Training data for this is available at
\cite{data}.

TF3 cooperates and defects with
various cycles depending on the opponent's actions. TF3 will mutually
cooperate with any strategy and only tolerates a few defections before
defecting for the rest of match. It is similar to but not exactly the same as
Fool Me Once, a strategy that cooperates until the opponent has defected twice
(not necessarily consecutively), and defects indefinitely thereafter. Though a
product of training with a Moran objective. It differs from TF1 and TF2
by lacking a handshake mechanism.

@@

Comparisons of \(x_1, x_{N/2}, x_{N-1}\) are shown in
Figure~\ref{fig:comparison_deterministic}. The points represent the simulated
values and the line shows the theoretical value. Note that these are all
deterministic strategies and show a perfect match up between the expected value
of (\ref{equ:fixation_probability}) and the actual Moran process for all
strategies pairs. Figure~\ref{fig:comparison_stochastic} shows the fixation probabilities for
stochastic strategies. These are no longer a good match which highlights the
weakness of assuming a given interaction between two IPD strategies can be
summarised with a set of utilities as shown in
(\ref{equ:payoff_matrix}). For any given pair of strategies it is possible to
obtain \(p_{i,i-1}, p_{i,i+1}, p_{ii}\) exactly (as opposed to the
approximations offered by (\ref{equ:p_up}), (\ref{equ:p_down}) and
(\ref{equ:p_stay}). Obtaining these requires particular analysis for a given
pair and can be quite a complex endeavour for stochastic strategies with long
memory: this is not necessary for the purposes of this work.
All data generated for this validation exercise can be found
at \cite{data}.

@@

As will be demonstrated in Section~\ref{sec:population_size} the results for
\(N=2\) differ from those of larger $N$. Hence these results do not concur with
the literature which suggests that zero Determinant strategies should be
effective for larger population sizes, but these analysis consider stationary
behaviour, while this work runs for a fixed number of rounds. \cite{stewart2013extortion}
The stationarity assumption allows for a deterministic payoff matrix
leading to the conclusions about zero determinant strategies in the space
of memory-one strategies that do not generalize to this context.

@@

Interestingly none of these strategies are stochastic: this is explained by
the need of strategies to have a steady hand when interacting with their own
kind. Acting stochastically increases the chance of friendly fire.
However it is possible to design a strategy with a stochastic or error-correcting
handshake that is an excellent resistor even in noisy environments \cite{Lee2015}.

@@

While TF1 and TF2 are competent invaders, the best invaders
in the study do not appear to employ strict handshakes, and are generally
cooperative strategies. TF3, which does not use a handshake, is a better invader
than TF1 and TF2 but not as good of a resistor. Nevertheless it was the result
of the same kind of training procceses and is a better combined invader-resistor
than the invaders that were trained previously to maximize payout.

@@

We did not attempt other objective functions that may serve to select for both
invasion and resistance better than training at a starting population of
$(N/2, N/2)$. Nevertheless our results suggest that there is not much room for
improvement. Any handshake more sophisticated than C necessarily involves
a defection. (A handshake consisting of a longer sequence of cooperations is
effectively a grudger.) For TF3 or EvolvedLookerUp1\_1\_1 to become better resistors
they need a longer or more strict handshake. But if this handshake involves
a defection then likely the invasion ability is diminished for $N > 2$: the top
invaders for larger $N$ are nice strategies that do not defect before their
opponents. This is because good invaders still need to cooperate with themselves,
and so in the absense of a handshake mechanism or knowledge of the population
distribution, a strategy must be generally cooperative. Aggressive strategies
are only effective invaders for the smallest $N$, dropping down dramatically
as the population size increases.

@@

Memory one strategies do not perform well in general in this study. There are
no memory one strategies in the top 5 performing strategies
for \(N>3\). This is due at least partly to their lack of ability to
recognise their opponents. More sophisticated strategies
prove to be high performers for invasion: these are infinite memory strategies
which have been trained using a number of reinforcement learning algorithms.
Interestingly they have been trained to perform well in tournaments and not
Moran processes which highlights the potentially for improvement.

@@

There are many opportunities build on this work. In particular, an analysis
of the effect of noise should offer insights regarding the stability of the findings,
particularly for the handshaking strategies. They may be less dominant for
larger amounts of noise since the handshaking mechanisms may become brittle.
There are many other variations to explore including populations with more
than one type, spatial structure, and mutation.

>>>>

\begin{enumerate}
 \item Zero determinant strategies are not particularly effective for $N > 2$
 \item Complex strategies can be effective, and in fact can naturally evolve
     through evolutionary processes to outperform designed strategies.
 \item Strong resistors specifically evolve or have a handshake mechanism.
 \item Strong invaders are generally cooperative strategies that do not defect
 first but retaliate to varying degrees of intensity against strategies that defect.
 \item Strategies evolved to maximize their total payoff can be strong invaders
 and achieve mutual cooperation with many other strategies.
\end{enumerate}

@@

Figure~\ref{fig:moran_process} shows a diagrammatic representation of the Moran
process, a stochastic birth death process on a finite population in which the
population size stays constant over time. Individuals are \textbf{selected}
according to a given fitness landscape. Once selected, the individual is
reproduced and similarly another individual is chosen to be removed from the
population. In some settings mutation is also considered but without mutation
(the case considered in this work) this process will arrive at an absorbing
state where the population is entirely made up of players of one strategy. The
probability with which a given strategy takes over a population is called the
\textit{fixation probability}. A more detailed analytic description of this is
given in Section~\ref{sec:validation}. In our simulations offspring do not
inherit any knowledge or history from parent replicants.

@@

The Moran process was initially introduced in \cite{Moran1957}. It has since
been used in a variety of settings including the understanding of the spread of
cooperative and non-cooperative behaviour such as cancer \cite{West2016} and the
emergence of cooperative behaviour in spatial topologies \cite{Nowak2017}.
However these works mainly consider non-sophisticated strategies. Some work has
looked at evolutionary stability of strategies within the Prisoner's Dilemma
\cite{Li2014} but this is not done in the more widely used setting of the Moran
process, rather in terms of infinite population stability. In \cite{Baek2016}
Moran processes are studied in a theoretical framework for a small subset of
strategies. The subset included memory one strategies; strategies that recall
the events of the previous round only.

Of particular interest are the Zero determinant strategies introduced in
\cite{Press2012}. It was argued in \cite{stewart2013extortion} that generous
ZD strategies are robust against invading strategies. However, in \cite{Lee2015},
a strategy using machine learning techniques was capable of resisting invasion
and also able to invade any memory one strategy. Recent work \cite{Hilbe2017}
has investigated the effect of memory length on strategy performance and the
emergence of cooperation but this is not done in a Moran process context and only
considers specific cases of memory 2 strategies. In \cite{Adami2013} it was
recognised that many zero determinant strategies do not fare well against
themselves. This is a disadvantage for the Moran process where the best
strategies cooperate well with other players using the same strategy.

@@

To carry out this large numerical experiment, \input{./tex/num_strategies}
strategies, listed (with their properties) in Appendix~\ref{app:list_of_players},
are used from the Axelrod library. There are
\input{./tex/num_stochastic.tex}stochastic and
\input{./tex/num_deterministic.tex}deterministic strategies. Their memory depth,
defined by the number of rounds of history used by the strategy each round, is
shown in Table~\ref{tbl:memory_depth_count}. The memory depth is infinite if the
strategy uses the entire history of play (whatever its length). For example, a
strategy that utilizes a handshaking mechanism where the opponent's actions on
the first few rounds of play determines the strategies subsequent behavior would
have infinite memory depth.

@@

Apart from the PSO Gambler strategy, which was trained using a particle swarm
optimisation algorithm, these strategies are trained with an evolutionary
algorithm that perturbs strategy parameters and optimizes the mean total score
against all other opponents \cite{affenzeller2009genetic}. They were trained to
win IPD tournaments by maximizing their mean total payoffs against a variety
of opponents. Variation is
introduced via mutation and crossover of parameters, and the best performing
strategies are carried to the next generation along with new variants. Similar
methods appear in the literature \cite{Ashlock2006}.

@@

Each matchup of players was run to fixation for the specified number of
repetitions to estimate the absorption probabilities. The trained algorithms
were run for fewer than 50 generations. Training data for this is available at
\cite{data}.

TF3 cooperates and defects with
various cycles depending on the opponent's actions. TF3 will mutually
cooperate with any strategy and only tolerates a few defections before
defecting for the rest of match. It is similar to but not exactly the same as
Fool Me Once, a strategy that cooperates until the opponent has defected twice
(not necessarily consecutively), and defects indefinitely thereafter. Though a
product of training with a Moran objective, it differs from TF1 and TF2
in that it lacks a handshake mechanism.

@@

Comparisons of \(x_1, x_{N/2}, x_{N-1}\) are shown in
Figure~\ref{fig:comparison_deterministic}. The points represent the simulated
values and the line shows the theoretical value. Note that these are all
deterministic strategies and show a perfect match between the expected value
of (\ref{equ:fixation_probability}) and the actual Moran process for all
strategy pairs. Figure~\ref{fig:comparison_stochastic} shows the fixation probabilities for
stochastic strategies. These are no longer a good match which highlights the
weakness of assuming a given interaction between two IPD strategies can be
summarised with a set of utilities as shown in
(\ref{equ:payoff_matrix}). For any given pair of strategies it is possible to
obtain \(p_{i,i-1}, p_{i,i+1}, p_{ii}\) exactly (as opposed to the
approximations offered by (\ref{equ:p_up}), (\ref{equ:p_down}) and
(\ref{equ:p_stay}). Obtaining these requires particular analysis for a given
pair and can be quite a complex endeavour for stochastic strategies with long
memory: this is not necessary for the purposes of this work.
All data generated for this validation exercise can be found
at \cite{data}.

@@

As will be demonstrated in Section~\ref{sec:population_size} the results for
\(N=2\) differ from those of larger $N$. Hence these results do not concur with
the literature which suggests that zero Determinant strategies should be
effective for larger population sizes, but these analyses consider stationary
behaviour, while this work runs for a fixed number of rounds. \cite{stewart2013extortion}
The stationarity assumption allows for a deterministic payoff matrix
leading to the conclusions about zero determinant strategies in the space
of memory-one strategies that do not generalize to this context.

@@

Interestingly none of these strategies is stochastic: this is explained by
the need of strategies to have a steady hand when interacting with their own
kind. Acting stochastically increases the chance of friendly fire.
However it is possible to design a strategy with a stochastic or error-correcting
handshake that is an excellent resistor even in noisy environments \cite{Lee2015}.

@@

While TF1 and TF2 are competent invaders, the best invaders
in the study do not appear to employ strict handshakes, and are generally
cooperative strategies. TF3, which does not use a handshake, is a better invader
than TF1 and TF2 but not as good a resistor. Nevertheless it was the result
of the same kind of training procceses and is a better combined invader-resistor
than the invaders that were trained previously to maximize payout.

@@

We did not attempt other objective functions that may serve to select for both
invasion and resistance better than training at a starting population of
$(N/2, N/2)$. Nevertheless our results suggest that there is not much room for
improvement. Any handshake more sophisticated than C necessarily involves
a defection. (A handshake consisting of a longer sequence of cooperations is
effectively a grudger.) For TF3 or EvolvedLookerUp1\_1\_1 to become better resistors
they need a longer or more strict handshake. But if this handshake involves
a defection then likely the invasion ability is diminished for $N > 2$: the top
invaders for larger $N$ are nice strategies that do not defect before their
opponents. This is because good invaders still need to cooperate with themselves,
and so in the absense of a handshake mechanism or knowledge of the population
distribution, a strategy must be generally cooperative. Aggressive strategies
are only effective invaders for the smallest $N$, dropping dramatically
as the population size increases.

@@

Memory one strategies do not perform well in general in this study. There are
no memory one strategies in the top 5 performing strategies
for \(N>3\). This is due at least partly to their lack of ability to
recognise their opponents. More sophisticated strategies
prove to be high performers for invasion: these are infinite memory strategies
which have been trained using a number of reinforcement learning algorithms.
Interestingly they have been trained to perform well in tournaments and not
Moran processes which highlights the potential for improvement.

@@

There are many opportunities to build on this work. In particular, an analysis
of the effect of noise should offer insights regarding the stability of the findings,
particularly for the handshaking strategies. They may be less dominant for
larger amounts of noise since the handshaking mechanisms may become brittle.
There are many other variations to explore including populations with more
than one type, spatial structure, and mutation.
